From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-23 Pi+/paper/finite.tex

\subsection{Permutation groups}

@@

\begin{definition}[$\longcox$]
  \begin{align*}
    \longcancel
     & : \forall n, l, r \to (l \append n \cons n \cons r) \longcox (l \append r)                                                                           \\
    \longswap
     & : \forall k, n, l, r \to (\suc[k] < n) \to (l \append n \cons k \cons r) \longcox (l \append k \cons n \append r)                                    \\
    \longbraid
     & : \forall n, l, r \to (l \append (n \downarrow 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downarrow 2 + k) \append r) \\
  \end{align*}
\end{definition}

@@

\begin{definition}[$\downf : (n : \Nat) \to {\Fin[n]} \to {\List[\Fin[\suc[k + n]]]}$]
  \begin{align*}
    n \downf \zero   & \defeq \nil                       \\
    n \downf \suc[k] & \defeq (k + n) \cons (n \downf k)
  \end{align*}
\end{definition}

@@

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item For all $l : \List[\Fin[n]]$, we have that $l \cox* \normf(l)$.
    \item $\normf$ is idempotent, that is, $\normf \comp \normf \htpy \normf$.
    \item $\normf$ splits, that is, we have ${\List[\Fin[n]]} \xrightarrow{s} \Sn[n] \xrightarrow{r} {\List[\Fin[n]]}$
          such that $s \comp r \htpy \normf$ and $r \comp s \htpy \idfunc_{\Sn[n]}$.
  \end{enumerate}
\end{proposition}

@@

\todo{Explain why the two definitions are equivalent?}
\note{immersion here?}

Suppose we have a permutation $\sigma$ on an $n$-element set $\{\el{0}, \el{1}, \el{2}, \el{3}, \el{4}\}$, we encode it
as follows. $\Lehmer[n]$ is a $n$-element tuple. At position $k$, we put the number of inversions of the element
$\el{k}$ in $p$, i.e. the number of elements smaller than $\el{k}$ occurring after $\el{k}$. As an example, consider the
following tabulated presentation of the permutation:

@@

Using this Lehmer encoding algorithm, we can now construct the equivalence between these types.

We define a type family $\FinExcept{n} : \Fin[n] \to \UU$ which picks out all elements in $\Fin[n]$ except the one
provided. Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence an $\hSet$.

@@

\begin{proposition}
  For all $n:\Nat$, \( \Lehmer[n] \eqv \Aut[\Fin[n]] \).
\end{proposition}

\begin{proof}
  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\zero]]$. For $n = \suc[m]$, we
  compute a chain of equivalences.
  \begin{gather*}
    \arraycolsep=0.5em\def\arraystretch{1.5}
    \begin{array}{rl}
           & \Lehmer[\zero]    \\
      \eqv & \unit             \\
      \eqv & \Aut[\Fin[\zero]] \\
    \end{array}
    \qquad\qquad
    \begin{array}{rl}
           & \Aut[\Fin[\suc[m]]]                                                               \\
      \eqv & \dsum{k : \Fin[\suc[m]]}{\FinExcept{\suc[m]}[\fzero] \eqv \FinExcept{\suc[m]}[k]} \\
      \eqv & \dsum{k : \Fin[\suc[m]]}{\FinExcept{\suc[m]}[\fzero] \eqv \Fin[m]}                \\
      \eqv & \dsum{k : \Fin[\suc[m]]}{\Fin[m] \eqv \Fin[m]}                                    \\
      \eqv & \Fin[\suc[m]] \times \Aut[\Fin[m]]                                                \\
      \eqv & \Fin[\suc[m]] \times \Lehmer[m]                                                   \\
      \eqv & \Lehmer[\suc[m]]
    \end{array}
  \end{gather*}
\end{proof}

We will now show how to generate a Lehmer code from a word in $\Sn$ and back.

\begin{definition}
  \[
    \encode{\List} : \List[\Fin[n]] \to \Lehmer[n]
  \]
\end{definition}

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item \( l_{1} \cox* l_{2} \to \encode{\List}(l_{1}) \id \encode{\List}(l_{1}) \)
  \end{enumerate}
\end{proposition}

\begin{definition}
  \begin{align*}
    \encodeSn & : \Sn[n] \to \Lehmer[n] \\
    \decodeSn & : \Lehmer[n] \to \Sn[n]
  \end{align*}
\end{definition}

\begin{proposition}
  For all $n : \Nat$, $(\encodeSn, \decodeSn)$ is an equivalence.
\end{proposition}

\begin{corollary}
  For all $n : \Nat$,
  \(
  \Sn \eqv \Lehmer[n] \eqv \Aut[\Fin[n]]
  \).
\end{corollary}

>>>>

\subsection{Presentation of permutation groups}

@@

\begin{definition}[$\longcox$]
  \begin{align*}
    \longcancel
     & : \forall n, l, r \to (l \append n \cons n \cons r) \longcox (l \append r)                                                                   \\
    \longswap
     & : \forall k, n, l, r \to (\suc[k] < n) \to (l \append n \cons k \cons r) \longcox (l \append k \cons n \append r)                            \\
    \longbraid
     & : \forall n, l, r \to (l \append (n \downf 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downf 2 + k) \append r) \\
  \end{align*}
\end{definition}

@@

\begin{definition}[$\downf : (n : \Nat) \to (k : \Nat) \to {\List[\Fin[\suc[k + n]]]}$]
  \begin{align*}
    n \downf \zero   & \defeq \nil                       \\
    n \downf \suc[k] & \defeq (k + n) \cons (n \downf k)
  \end{align*}
\end{definition}

@@

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item For all $l : \List[\Fin[n]]$, we have that $l \cox* \normf(l)$.
    \item $\normf$ is idempotent, that is, $\normf \comp \normf \htpy \normf$.
    \item $\normf$ splits, that is, we have ${\List[\Fin[n]]} \xrightarrow{s} \Sn[n] \xrightarrow{r} {\List[\Fin[n]]}$
          such that $s \comp r \htpy \normf$ and $r \comp s \htpy \idfunc_{\Sn[n]}$.
    \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \).
  \end{enumerate}
\end{proposition}

@@

\todo{Explain why the two definitions are equivalent?}

Suppose we have a permutation $\sigma$ on an $n$-element set $\{\el{0}, \el{1}, \el{2}, \el{3}, \el{4}\}$, we encode it
as follows. $\Lehmer[n]$ is a $n$-element tuple. Then, At position $k$, we put the number of inversions of the element
$\el{k}$ in $p$, i.e. the number of elements smaller than $\el{k}$ occurring after $\el{k}$. As an example, consider the
following tabulated presentation of the permutation:

@@

Writing formally, to turn a code into a permutation written in the Coxeter presentation, we use $\immersion$:

\begin{definition}[$\immersion : (n : \Nat) \to {\Lehmer[n]} \to {\List[\Fin[n]]}$]
  \begin{align*}
    \immersion(\zero, \zero)    & \defeq \nil                                              \\
    \immersion(\suc[n], (r, l)) & \defeq \immersion(n, l) \append ((\suc[n] - r) \downf r)
  \end{align*}
\end{definition}

As described above, the number $r$ at the position $k$ in the tuple describes how many inversions the element $\el{k}$
has. Thus, we need to perform $r$ many adjacent transpositions to get to the desired position. The function that
constructs a sequence of $r$ adjacent transpositions $[\suc[n], \suc[n] - 1, \suc[n] - 2, \ldots \suc[n] - r]$ is
$(\suc[n] - r) \downf r$.

We can show that the function $\immersion$ gives an equivalence betweeen $\Lehmer[n]$ and $\im{\normf}$.

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item For any Lehmer code $c$, $\immersion(c)$ is a normal form with resepct to $\cox*$, that is, $\immersion(c)$ is
          in $\im{\normf}$.
    \item Any element of $\im{\normf}$ can be constructed from a unique Lehmer code by $\immersion$, that is, the fibers
          of $\immersion: \Lehmer[n] \to {\im{\normf}}$ are contractible.
  \end{enumerate}
  Therefore, there is an equivalence between $\Lehmer[n]$ and $\im{\normf}$.
\end{proposition}

We can also establish an equivalence between $\im{\normf}$ and $\Sn$, which gives the following.

\begin{proposition}
  For all $n : \Nat$, \( \Sn \eqv \im{\normf} \eqv \Lehmer[n] \).
\end{proposition}

\subsection{Running Lehmer codes}

Finally, it is time to complete our goal of characterising the permutation groups. Having produced a Lehmer code by
normalising words in $\Sn$, we need to run it to produce a concrete bijection of finite sets, and, given a bijection
between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$.

To start, we define a type family $\FinExcept{n} : \Fin[n] \to \UU$ which picks out all elements in $\Fin[n]$ except the
one provided. Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence an $\hSet$.

@@

\begin{proposition}
  For all $n:\Nat$, \( \Lehmer[n] \eqv \Aut[\Fin[\suc[n]]] \).
\end{proposition}

\begin{proof}
  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\suc[\zero]]]$. For $n = \suc[m]$,
  we compute a chain of equivalences.
  \begin{gather*}
    \arraycolsep=0.5em\def\arraystretch{1.5}
    \begin{array}{rl}
           & \Lehmer[\zero]          \\
      \eqv & \unit                   \\
      \eqv & \Aut[\Fin[\suc[\zero]]] \\
    \end{array}
    \qquad\qquad
    \begin{array}{rl}
           & \Aut[\Fin[\suc[\suc[m]]]]                                                                     \\
      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \FinExcept{\suc[m]}[k]} \\
      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \Fin[\suc[m]]}          \\
      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                              \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Aut[\Fin[\suc[m]]]                                                \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Lehmer[m]                                                         \\
      \eqv & \Lehmer[\suc[m]]
    \end{array}
  \end{gather*}
\end{proof}

\begin{corollary}
  For all $n : \Nat$,
  \(
  \Sn \eqv \Lehmer[n] \eqv \Aut[\Fin[\suc[n]]]
  \).
\end{corollary}
